---
title: "A1_KatieLee"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Table of Contents

## 1. Introduction


### Required libraries and functions:
```{r message=FALSE}
library(knitr)
library(GEOquery)
library(readr)
library(biomaRt)
library(dplyr)

# Prints the first n number of dataframe x, with an optional caption
kable_head <- function(x, n = 5, caption = NULL) {
  knitr::kable(utils::head(x, n), caption = caption)
}
```

## 2. Preparing and Cleaning the Dataset

### 2.1 Downloading the Dataset

```{r}
# Downloads data into a new or existing directory, default to data
fetch_geo_supp <- function(gse, destdir = "data") {
  dir.create(destdir, showWarnings = FALSE, recursive = TRUE)
  GEOquery::getGEOSuppFiles(GEO = gse, baseDir = destdir, makeDirectory = TRUE)
  invisible(file.path(destdir, gse))
}

gse <- "GSE226507"
geo_path <- file.path("data", gse)

# Download the dataset if it doesn't already exist
if (!dir.exists(geo_path)) {
  fetch_geo_supp(gse = gse)
}

# GEO file contains only one file
geo_file <- list.files(file.path("data", gse), full.names = TRUE)[1]
```

### 2.2 Raw Data Table
```{r}
# Reading the dataset
geo_df_raw <- readr::read_tsv(geo_file, show_col_types = FALSE)

kable_head(geo_df_raw, caption=paste(gse, ": Raw Data Table Preview"))
```
### 2.3 Remove Extra Annotations
The data set includes extra annotations, such as gene length, start, end, description, etc. To simplify the downstream analyses, these annotations were removed.
```{r}
# List of all the samples in the data set
sample_cols <- c("WC1","WC2","WC3","WSD1","WSD2","WSD3",
                 "KC1","KC2","KC3","KSD1","KSD2","KSD3")

# Remove everything other than the sample columns
geo_df <- geo_df_raw[, c("gene_id", sample_cols)]
```
### 2.4 Checking Data Quality
I checked the dataset for any duplicates in the sample or genes. I also checked for missing values.
```{r}
any(duplicated(geo_df$gene_id)) # True if duplicate genes
any(duplicated(colnames(geo_df))) # True if duplicate samples

sum(is.na(geo_df))
```
This dataset does not contain any duplicates or missing values, so I proceeded with the downstream analyses.

### 2.5 Libary Size
I calculated the library size for each sample. To do this, I calculated total counts across all genes for a sample. 
```{r}
library_sizes <- colSums(geo_df[ , -1], na.rm = TRUE) # geo_df[, -1] uses only the sample count columns

library_sizes_df <- data.frame(
  sample = names(library_sizes),
  library_size = library_sizes
)

# Removes duplicate row names (sample col already exists)
rownames(library_sizes_df) <- NULL

kable_head(library_sizes_df, n=12, caption=paste(gse, ": Library Size Table"))
```

I calculated the summary statistics of gene expressions for each sample. 
```{r}
sample_stats <- data.frame(
  sample = colnames(geo_df)[-1],
  mean = colMeans(geo_df[, -1]),
  median = apply(geo_df[, -1], 2, median),
  sd_expression = apply(geo_df[, -1], 2, sd)
)
rownames(sample_stats) <- NULL

kable_head(sample_stats, 12, caption="Summary Statistics for Each Sample")
```
### 2.5 HUGO Gene Symbol Mapped Table
The chosen GSE file does not have any version suffixes, so it does not need to be removed. 
```{r}
# Getting Gene ID Column
id_col <- names(geo_df)[1]
ids <- geo_df[[1]] |> as.character()

# Getting ensembl mart and dataset for all the Gene IDs
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
```

```{r}
# Mapping the Gene ID to HUGO Gene Symbol
ensembl_ids <- unique(ids)

map <- getBM(
    attributes = c("ensembl_gene_id", "hgnc_symbol"),
    filters = "ensembl_gene_id",
    values = ensembl_ids,
    mart = ensembl
  )

# Adding the mapped HGNC symbol into GSE dataframe
df_mapped <- geo_df %>%
  left_join(map, by=c("gene_id" = "ensembl_gene_id")) %>%
  dplyr::select(gene_id, hgnc_symbol, everything())
  
kable_head(df_mapped, caption=paste(gse, ": Mapped Table Preview"))
```

## 3. Normalization of Data


## 4. Differential Gene Expression

