---
title: "A1_KatieLee"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Table of Contents

## 1. Introduction


### Required libraries and functions:
```{r message=FALSE}
library(knitr)
library(GEOquery)
library(readr)
library(biomaRt)
library(dplyr)
library(tidyverse)
library(edgeR)

# Prints the first n number of dataframe x, with an optional caption
kable_head <- function(x, n = 5, caption = NULL) {
  knitr::kable(utils::head(x, n), caption = caption)
}
```

## 2. Preparing and Cleaning the Dataset

### 2.1 Downloading the Dataset

```{r}
# Downloads data into a new or existing directory, default to data
fetch_geo_supp <- function(gse, destdir = "data") {
  dir.create(destdir, showWarnings = FALSE, recursive = TRUE)
  GEOquery::getGEOSuppFiles(GEO = gse, baseDir = destdir, makeDirectory = TRUE)
  invisible(file.path(destdir, gse))
}

gse <- "GSE226507"
geo_path <- file.path("data", gse)

# Download the dataset if it doesn't already exist
if (!dir.exists(geo_path)) {
  fetch_geo_supp(gse = gse)
}

# GEO file contains only one file
geo_file <- list.files(file.path("data", gse), full.names = TRUE)[1]
```

### 2.2 Raw Data Table
```{r}
# Reading the dataset
geo_df_raw <- readr::read_tsv(geo_file, show_col_types = FALSE)

kable_head(geo_df_raw, caption=paste(gse, ": Raw Data Table Preview"))
```
### 2.3 Remove Extra Annotations
The data set includes extra annotations, such as gene length, start, end, description, etc. To simplify the downstream analyses, these annotations were removed.
```{r}
# List of all the samples in the data set
sample_cols <- c("WC1","WC2","WC3","WSD1","WSD2","WSD3",
                 "KC1","KC2","KC3","KSD1","KSD2","KSD3")

# Remove everything other than the sample columns
geo_df <- geo_df_raw[, c("gene_id", sample_cols)]
```
### 2.4 Checking Data Quality
I checked the dataset for any duplicates in the sample or genes. I also checked for missing values.
```{r}
any(duplicated(geo_df$gene_id)) # True if duplicate genes
any(duplicated(colnames(geo_df))) # True if duplicate samples

sum(is.na(geo_df))
```
This dataset does not contain any duplicates or missing values, so I proceeded with the downstream analyses.

### 2.5 Libary Size
I calculated the library size for each sample. To do this, I calculated total counts across all genes for a sample. 
```{r}
library_sizes <- colSums(geo_df[ , -1], na.rm = TRUE) # geo_df[, -1] uses only the sample count columns

library_sizes_df <- data.frame(
  sample = names(library_sizes),
  library_size = library_sizes
)

# Removes duplicate row names (sample col already exists)
rownames(library_sizes_df) <- NULL

kable_head(library_sizes_df, n=12, caption=paste(gse, ": Library Size Table"))
```

I calculated the summary statistics of gene expressions for each sample. 
```{r}
sample_stats <- data.frame(
  sample = colnames(geo_df)[-1],
  mean = colMeans(geo_df[, -1]),
  median = apply(geo_df[, -1], 2, median),
  sd_expression = apply(geo_df[, -1], 2, sd)
)
rownames(sample_stats) <- NULL

kable_head(sample_stats, 12, caption="Summary Statistics for Each Sample")
```
### 2.5 HUGO Gene Symbol Mapped Table
The chosen GSE file does not have any version suffixes, so it does not need to be removed. 
```{r}
# Getting Gene ID Column
id_col <- names(geo_df)[1]
ids <- geo_df[[1]] |> as.character()

# Getting ensembl mart and dataset for all the Gene IDs
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
```

```{r}
# Mapping the Gene ID to HUGO Gene Symbol
ensembl_ids <- unique(ids)

map <- getBM(
    attributes = c("ensembl_gene_id", "hgnc_symbol"),
    filters = "ensembl_gene_id",
    values = ensembl_ids,
    mart = ensembl
  )

# Adding the mapped HGNC symbol into GSE dataframe
df_mapped <- geo_df %>%
  left_join(map, by=c("gene_id" = "ensembl_gene_id")) %>%
  dplyr::select(gene_id, hgnc_symbol, everything())
```
### 2.6 Cleaning Mapped Data
Many genes did not map to a HUGO symbol, so they were removed from the analysis. 
```{r}
df_mapped <- df_mapped %>%
  filter(!is.na(hgnc_symbol), hgnc_symbol != "")
```

Some genes also mapped to the same HUGO symbol, so I used the mean expression of these genes. No genes had multiple HUGO symbols.
```{r}
# Checking if there are any genes that map to the same HUGO symbol
if (any(duplicated(df_mapped$hgnc_symbol))) {
  df_mapped <- df_mapped %>%
    group_by(hgnc_symbol) %>%
    summarise(across(where(is.numeric), mean), .groups = "drop") 
  # finds mean expression and drops the grouping
}
```
I cleaned the dataframe, so it has the HUGO symbol as rownames.
```{r}
df_mapped <- as.data.frame(df_mapped)
rownames(df_mapped) <- df_mapped$hgnc_symbol
df_mapped$hgnc_symbol <- NULL

kable_head(df_mapped, n=10, caption="Expression Data of Mapped HUGO Symbols for GSE226507")
```
## 3. Normalization of Data
### 3.1 Separating Sample Groups
```{r}
# All the treatment groups of this experiment
treatments <- c("WC", "WSD", "KC", "KSD")
# Gets the columns with the sample expression data
samples <- colnames(df_mapped)

# Groups the sample to the treatment group it starts with
samples_to_groups <- data.frame(
  sample=samples,
  group=sapply(samples, function(x) treatments[startsWith(x, treatments)])
)

rownames(samples_to_groups) <- NULL

samples_to_groups
```

### 3.2 Filtering Lowly Expressed Genes
```{r}
group <- factor(samples_to_groups$group)
design <- model.matrix(~ 0 + group, data = samples_to_groups)
to_keep <- edgeR::filterByExpr(df_mapped, min.count = 3, design = design)
df_mapped_filtered <- df_mapped[to_keep, ]
```

### 3.4 TMM Normalization
```{r}
dge <- DGEList(counts = df_mapped_filtered)
dge_filtered <- dge[filterByExpr(dge),]
dge_filtered  <- calcNormFactors(dge_filtered , method = "TMM")
df_normalized <- cpm(dge_filtered , log = FALSE, prior.count = 1)

kable_head(df_normalized, n=10, caption="TMM-normalized logCPM expression matrix of GSE226507")
```
### 3.5 Comparison of Pre- and Post Normalization
```{r}
plot_box <- function(mat, metadata, group, main = "", ylab = "log2(counts+1)") {
  df <- as.data.frame(mat)
  df_long <- df |>
    mutate(gene = rownames(df)) |>
    pivot_longer(-gene, names_to = "sample", values_to = "value") |>
    mutate(value = log2(value + 1)) |>
    left_join(metadata, by = "sample")

  ggplot(df_long, aes(x = sample, y = value, fill= .data[[group]])) +
    geom_boxplot(outlier.size = 0.2) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    labs(title = main, x = "Sample", y = ylab, fill=group)
}

plot_density <- function(mat, metadata, group, main = "", xlab="log2(counts+1)") {
  df <- as.data.frame(mat)
  df_long <- df |>
    mutate(gene = rownames(df)) |>
    pivot_longer(-gene, names_to = "sample", values_to = "value") |>
    mutate(value = log2(value + 1)) |>
    left_join(metadata, by = "sample")

  ggplot(df_long, aes(x = value, colour = .data[[group]])) +
    geom_density() +
    theme_bw() +
    labs(title = main, x = xlab, y = "Density", colour = group)
}

plot_box(df_normalized, samples_to_groups, "group", main="Boxplot of normalized expression for GSE226507", ylab="log2(CPM + 1)")
plot_density(df_normalized, samples_to_groups, "group", main="Density plot of normalized expression for GSE226507", xlab="log2(CPM + 1)")

plot_box(df_mapped_filtered, samples_to_groups, "group", main="Boxplot of raw expression for GSE226507")
plot_density(df_mapped_filtered, samples_to_groups, "group", main="Density plot of raw expression for GSE226507")

```
## 4. Differential Gene Expression

## 5. Conclusion
To summarize my findings and report, I will answer the following questions:
1. Why is the dataset of interest to you?
2. What are the control and test conditions of the dataset?
3. How many samples in each of the conditions of your dataset?
4. Were there expression values that were not unique for specific genes? How did you handle these?
5. Were there expression values that could not be mapped to current HUGO symbols?
6. Were there any outliers in your dataset? How were they handled in the originating paper? How many outliers were removed?
7. How did you handle replicates?
8. What is the final coverage of your dataset?

## 6. References

