# Exercises

These are intended to be done **after** completing the worked examples.

## Exercise 1 — https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE119732

Using **GSE119732**, confirm whether the ID column contains Ensembl IDs with version suffixes.

1. Extract the first 20 IDs.
2. Count how many contain a `.`.
3. Create a new column with versions stripped.
4. Map the identifiers to HGNC symbols.
```{r message=FALSE}
library(knitr)
library(GEOquery)
library(readr)
library(biomaRt)
library(dplyr)
```

```{r message=FALSE}
fetch_geo_supp <- function(gse, destdir = "data") {
  dir.create(destdir, showWarnings = FALSE, recursive = TRUE)
  GEOquery::getGEOSuppFiles(GEO = gse, baseDir = destdir, makeDirectory = TRUE)
  invisible(file.path(destdir, gse))
}

# 1. Extract first 20 IDs
gse <- "GSE119732"
geo_path <- file.path("data", gse)

fetch_geo_supp(gse = gse)
file <- list.files(file.path("data", gse), full.names = TRUE)
df <- readr::read_tsv(file, show_col_types = FALSE)

df <- df[1:20, ]
head(df)
```

```{r}
# 2. Count how many contain a `.`
sum(grepl("\\.[0-9]+$", df$gene_id))
```
```{r}
# 3. Create a new column with versions stripped.
strip_ensembl_version <- function(x) sub("\\..*$", "", x)

id_col <- names(df)[1]
ids <- df[[1]] |> as.character()
if (sum(grepl("\\.[0-9]+$", ids)) >= 1) {
  ids <- strip_ensembl_version(ids)
}
df$versions_stripped <- ids
head(df)
```
```{r}
# 4. Map the identifiers to HGNC symbols.
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
map <- getBM(
    attributes = c("ensembl_gene_id", "hgnc_symbol"),
    filters = "ensembl_gene_id",
    values = ids,
    mart = ensembl
  )

map$ensembl_gene_id <- as.character(map$ensembl_gene_id)

df_mapped <- df %>%
  left_join(map, by=c("versions_stripped" = "ensembl_gene_id")) %>%
  dplyr::select(gene_id, hgnc_symbol, everything())

```
```{r}
df_mapped
```
## Exercise 2 — https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE122380

Using **GSE122380**, confirm whether the ID column contains Ensembl IDs with version suffixes.

1. Extract the first 20 IDs.
2. Create a new column with versions stripped.
3. Map the identifiers to HGNC symbols.
4. What is different about this file? 

The file is formatted weirdly and the columns are not matched to the samples.

```{r message=FALSE}
fetch_geo_supp <- function(gse, destdir = "data") {
  dir.create(destdir, showWarnings = FALSE, recursive = TRUE)
  GEOquery::getGEOSuppFiles(GEO = gse, baseDir = destdir, makeDirectory = TRUE)
  invisible(file.path(destdir, gse))
}

# 1. Extract first 20 IDs
gse <- "GSE122380"
geo_path <- file.path("data", gse)

fetch_geo_supp(gse = gse)
file <- list.files(file.path("data", gse), full.names = TRUE)[1]
df <- readr::read_tsv(file, show_col_types = FALSE)

df <- df[1:20, ]
head(df)
```

```{r}
# 3. Create a new column with versions stripped.
strip_ensembl_version <- function(x) sub("\\..*$", "", x)

id_col <- names(df)[1]
ids <- df[[1]] |> as.character()
if (sum(grepl("\\.[0-9]+$", ids)) >= 1) {
  ids <- strip_ensembl_version(ids)
}
df$versions_stripped <- ids
head(df)
```

```{r}
# 4. Map the identifiers to HGNC symbols.
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
map <- getBM(
    attributes = c("ensembl_gene_id", "hgnc_symbol"),
    filters = "ensembl_gene_id",
    values = ids,
    mart = ensembl
  )

map$ensembl_gene_id <- as.character(map$ensembl_gene_id)

df_mapped <- df %>%
  left_join(map, by=c("versions_stripped" = "ensembl_gene_id")) %>%
  dplyr::select(Gene_id, hgnc_symbol, everything())


df_mapped
```

## Exercise 3 - 
 
Can you use the worked example to process the above two GEO records?  How?

If the worked example contains genes that are in the two GEO records, we can use that data to map the gene ids to HUGO symbol.