# Exercises
These are intended to be done **after** completing the worked examples.

## Exercise 1 — https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE119732

Using **GSE119732**, 

  1. Map the distribution of the samples as a box plot and a density plot using only the raw data.
  2.  Add colours to the plot separating the different variables in the model
  3.  Filter out lowly expressed genes and re plot the distributions.
```{r}
fetch_geo_supp <- function(gse, destdir = "data") {
  dir.create(destdir, showWarnings = FALSE, recursive = TRUE)
  GEOquery::getGEOSuppFiles(GEO = gse, baseDir = destdir, makeDirectory = TRUE)
  invisible(file.path(destdir, gse))
}

gse <- "GSE119732"
geo_path <- file.path("data", gse)

# Download the dataset if it doesn't already exist
if (!dir.exists(geo_path)) {
  fetch_geo_supp(gse = gse)
}

# GEO file contains only one file
geo_file <- list.files(file.path("data", gse), full.names = TRUE)[1]
geo_df_raw <- readr::read_tsv(geo_file, show_col_types = FALSE)
head(geo_df_raw)
```
```{r}
# 1. Map the distribution of the samples as a box plot and a density plot using only the raw data.
library(tidyverse)

plot_box <- function(mat, main = "", ylab = "log2(counts+1)") {
  df <- as.data.frame(mat)
  df_long <- df |>
    mutate(gene = rownames(df)) |>
    pivot_longer(-gene, names_to = "sample", values_to = "value") |>
    mutate(value = log2(value + 1))

  ggplot(df_long, aes(x = sample, y = value)) +
    geom_boxplot(outlier.size = 0.2) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    labs(title = main, x = "Sample", y = ylab)
}

plot_density <- function(mat, main = "") {
  df <- as.data.frame(mat)
  df_long <- df |>
    mutate(gene = rownames(df)) |>
    pivot_longer(-gene, names_to = "sample", values_to = "value") |>
    mutate(value = log2(value + 1))

  ggplot(df_long, aes(x = value, colour = sample)) +
    geom_density() +
    theme_bw() +
    labs(title = main, x = "log2(counts+1)", y = "Density") +
    guides(colour = "none")
}

plot_box(geo_df_raw[,2:ncol(geo_df_raw)])
plot_density(geo_df_raw[,2:ncol(geo_df_raw)])
```
```{r}
# 2.  Add colours to the plot separating the different variables in the model

plot_box_colour <- function(mat, metadata, group, main = "", ylab = "log2(counts+1)") {
  df <- as.data.frame(mat)
  df_long <- df |>
    mutate(gene = rownames(df)) |>
    pivot_longer(-gene, names_to = "sample", values_to = "value") |>
    mutate(value = log2(value + 1)) |>
    left_join(metadata, by = "sample")

  ggplot(df_long, aes(x = sample, y = value, fill= .data[[group]])) +
    geom_boxplot(outlier.size = 0.2) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    labs(title = main, x = "Sample", y = ylab, fill=group)
}

plot_density_colour <- function(mat, metadata, group, main = "") {
  df <- as.data.frame(mat)
  df_long <- df |>
    mutate(gene = rownames(df)) |>
    pivot_longer(-gene, names_to = "sample", values_to = "value") |>
    mutate(value = log2(value + 1)) |>
    left_join(metadata, by = "sample")

  ggplot(df_long, aes(x = value, colour = .data[[group]])) +
    geom_density() +
    theme_bw() +
    labs(title = main, x = "log2(counts+1)", y = "Density", colour = group)
}

count_mat <- as.data.frame(geo_df_raw[, 2:ncol(geo_df_raw)])
rownames(count_mat) <- geo_df_raw[[1]]

sample_info <- data.frame(
  sample = colnames(count_mat)
) |>
  mutate(group = substr(sample, 1, 1))

plot_box_colour(count_mat, sample_info, "group")
plot_density_colour(count_mat, sample_info, "group")

```
```{r}
#   3.  Filter out lowly expressed genes and re plot the distributions.
to_keep <- edgeR::filterByExpr(count_mat,min.count = 3)
count_mat_filtered <- count_mat[to_keep,]

plot_box_colour(count_mat_filtered, sample_info, "group")
plot_density_colour(count_mat_filtered, sample_info, "group")
```

## Exercise 2 — https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE122380

Using **GSE122380**, confirm whether the ID column contains Ensembl IDs with version suffixes.

 1. Map the distribution of the samples as a box plot and a density plot using only the raw data.
  2.  Add colours to the plot separating the different variables in the model
  3.  Filter out lowly expressed genes and re plot the distributions.
  
```{r}
# 1. Map the distribution of the samples as a box plot and a density plot using only the raw data.

gse <- "GSE122380"
geo_path <- file.path("data", gse)

# Download the dataset if it doesn't already exist
if (!dir.exists(geo_path)) {
  fetch_geo_supp(gse = gse)
}

# GEO file contains only one file
geo_file <- list.files(file.path("data", gse), full.names = TRUE)[2]
geo_df_raw <- readr::read_tsv(geo_file, show_col_types = FALSE)
head(geo_df_raw)

plot_box(geo_df_raw[,2:ncol(geo_df_raw)])
plot_density(geo_df_raw[,2:ncol(geo_df_raw)])
```
```{r}
# 2.  Add colours to the plot separating the different variables in the model

count_mat <- as.data.frame(geo_df_raw[, 2:ncol(geo_df_raw)])
rownames(count_mat) <- geo_df_raw[[1]]

sample_info <- data.frame(
  sample = colnames(count_mat)
) |>
  mutate(group = substr(sample, 1, 1))

plot_box_colour(count_mat, sample_info, "group")
plot_density_colour(count_mat, sample_info, "group")

# 3.  Filter out lowly expressed genes and re plot the distributions.
to_keep <- edgeR::filterByExpr(count_mat,min.count = 3)
count_mat_filtered <- count_mat[to_keep,]

plot_box_colour(count_mat_filtered, sample_info, "group")
plot_density_colour(count_mat_filtered, sample_info, "group")
```

## Exercise 3 - 
 
Can you use the worked example to process the above two GEO records?  How?